---
title: "Mounting Filesystem Data"
format: html
toc: true
---

## The virtual filesystem

The [Emscripten filesystem API](https://emscripten.org/docs/api_reference/Filesystem-API.html) provides a Unix-like virtual filesystem for the WebAssembly (Wasm) R process running in webR. This virtual filesystem has the ability to [mount](https://emscripten.org/docs/api_reference/Filesystem-API.html#FS.mount) filesystem images or host directories so that the associated file and directory data is accessible to the Wasm R process.

Mounting images and directories in this way gives the Wasm R process access to arbitrary external data, potentially including datasets, scripts, or R packages [pre-compiled for WebAssembly](building.qmd).

Emscripten's API allows for several types of virtual filesystem, depending on the execution environment. The following filesystems are available for use with webR:

| Filesystem | Description | Web Browser | Node.js |
|------|-----|------|------|
| `WORKERFS` | Mount Emscripten filesystem images. | &#x2705; | &#x2705; |
| `NODEFS` | Mount existing host directories. | &#x274C; | &#x2705; |
| `IDBFS` | Browser-based persistent storage using the [IndexedDB API](https://developer.mozilla.org/en-US/docs/Web/API/IndexedDB_API). | &#x2705;[^idbfs] | &#x274C; |
| `DRIVEFS` | Mount a [JupyterLite](https://jupyterlite.readthedocs.io/en/stable/) DriveFS filesystem. | &#x2705;[^lite] | &#x274C; |

[^idbfs]: Using the `PostMessage` [communication channel](communication.qmd) only.
[^lite]: When webR is running as a [JupyterLite kernel](https://github.com/r-wasm/jupyterlite-webr-kernel) only.

## Filesystem images

Filesystem images are pre-prepared files containing a collection of files and associated metadata. The `WORKERFS` filesystem type can be used to efficiently make the contents of a filesystem image available to the WebAssembly R process.

### Emscripten's `file_packager` tool

The [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool, provided by Emscripten, takes in a directory structure as input and produces a webR compatible filesystem image as output. The [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool may be invoked from R using the [rwasm](https://r-wasm.github.io/rwasm/) R package:

```{r eval=FALSE}
> rwasm::file_packager("./input", out_dir = ".", out_name = "output")
```

It can also be invoked directly using its CLI^[See the [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) Emscripten documentation for details. ], if you prefer:

```bash
$ file_packager output.data --preload ./input@/ \
    --separate-metadata --js-output=output.js
```

In the above examples, the files in the directory `./input` are packaged and an output filesystem image is created^[When using the `file_packager` CLI, a third file named `output.js` will also be created. If you only plan to mount the image using webR, this file may be discarded.] consisting of a data file, `output.data`, and a metadata file, `output.js.metadata`.

To prepare for mounting the filesystem image with webR, ensure that both files have the same basename (in this example, `output`). The resulting URLs or relative paths for the two files should differ only by the file extension.

#### Compression

Filesystem image `.data` files may optionally be `gzip` compressed prior to deployment. The file extension for compressed filesystem images should be `.data.gz`, and compression should be indicated by setting the property `gzip: true` on the metadata JSON stored in the `.js.metadata` file.

### Process archives with the `rwasm` package

Archives in `.tar` format, optionally gzip compressed as `.tar.gz` or `.tgz` files, can also be used as filesystem images by pre-processing the `.tar` archive using the [rwasm](https://r-wasm.github.io/rwasm/) R package. The `rwasm::add_tar_index()` function reads the archive contents and appends the required filesystem metadata to the end of the `.tar` archive data in a way that is understood by webR.

```{r eval=FALSE}
> rwasm::add_tar_index("./path/to/archive.tar.gz")
```

Once processed by the `rwasm` R package, the archive can be deployed and used directly as a filesystem image.

## Mounting a filesystem image

When running in a web browser, the [`webr::mount()`](api/r.qmd#mount) function downloads and mounts a filesystem image from a URL source, using the `WORKERFS` filesystem type.

```{r eval=FALSE}
webr::mount(
  mountpoint = "/data",
  source = "https://example.com/output.data"
)
```

Filesystem images should be deployed to static file hosting^[e.g. GitHub Pages, Netlify, AWS S3, etc.] and the resulting URL provided as the source argument. The image will be mounted in the virtual filesystem under the path given by the `mountpoint` argument. If the `mountpoint` directory does not exist, it will be created prior to mounting.

When running under Node.js, the source may also be provided as a relative path to a filesystem image on disk.

### JavaScript API

WebR's JavaScript API includes the [`WebR.FS.mount()`](api/js/classes/WebR.WebR.md#fs) function, a thin wrapper around Emscripten's own [`FS.mount()`](https://emscripten.org/docs/api_reference/Filesystem-API.html#FS.mount). The JavaScript API provides more flexibility but requires a little more set up, including creating the `mountpoint` directory if it does not already exist.

The filesystem type should be provided as a `string`, with the `options` argument of type [`FSMountOptions`](api/js/modules/WebR.md#fsmountoptions). The filesystem image data should be provided either as a JavaScript `Blob` object or an `ArrayBuffer`-like object, and the metadata provided as a JavaScript object that has been deserialised from the underlying JSON content.

::: {.panel-tabset}
## JavaScript

``` javascript
// Create mountpoint
await webR.FS.mkdir('/data')

// Download image data
const data = await fetch('https://example.com/output.data');
const metadata = await fetch('https://example.com/output.js.metadata');

// Mount image data
const options = {
  packages: [{
    blob: await data.blob(),
    metadata: await metadata.json(),
  }],
}
await webR.FS.mount("WORKERFS", options, '/data');
```

## TypeScript

``` typescript
import { FSMountOptions } from 'webr';

// Create mountpoint
await webR.FS.mkdir('/data')

// Download image data
const data = await fetch('https://example.com/output.data');
const metadata = await fetch('https://example.com/output.js.metadata');

// Mount image data
const options: FSMountOptions = {
  packages: [{
    blob: await data.blob(),
    metadata: await metadata.json(),
  }],
}
await webR.FS.mount("WORKERFS", options, '/data');
```

:::

See the [Emscripten `FS.mount()` documentation](https://emscripten.org/docs/api_reference/Filesystem-API.html#FS.mount) for further details about the structure of the `options` argument.

## Mount an existing host directory

The `NODEFS` filesystem type maps directories that exist on the host machine so that they are accessible in the WebAssembly process.

::: callout-warning
`NODEFS` is only available when running webR under Node.js.
:::

To mount the directory `./extra` on the virtual filesystem at `/data`, use either the JavaScript or R mount API with the filesystem type set to `"NODEFS"`.

::: {.panel-tabset}
## JavaScript

``` javascript
await webR.FS.mkdir('/data')
await webR.FS.mount('NODEFS', { root: './extra' }, '/data');
```

## R
```{r eval=FALSE}
webr::mount(
  mountpoint = "/data",
  source = "./extra",
  type = "NODEFS"
)
```

:::

## IndexedDB Filesystem Storage

When using webR in a web browser, an [IndexedDB](https://developer.mozilla.org/en-US/docs/Web/API/IndexedDB_API)-based persistent storage space can be mounted using the `IDBFS` filesystem type.

::: {.callout-warning}

Due to the way webR blocks for input in the worker thread, the `IDBFS` filesystem type **does not work** when using the `SharedArrayBuffer` communication channel. WebR must be configured to use the `PostMessage` communication channel to use `IDBFS` persistent storage.

:::

### Mounting

First, create a directory to contain the IndexedDB filesystem, then use either the JavaScript or R mount API with type `"IDBFS"`.

::: {.panel-tabset}
## JavaScript

``` javascript
await webR.FS.mkdir('/data');
await webR.FS.mount('IDBFS', {}, '/data');
await webR.FS.syncfs(true);
```

## R
```{r eval=FALSE}
dir.create("/data")
webr::mount(mountpoint = "/data", type = "IDBFS")
webr::syncfs(TRUE)
```

:::

After mounting the filesystem using [`mount()`](api/r.html#mount), the [`syncfs()`](api/r.html#syncfs) function should been invoked with its `populate` argument set to `true`. This extra step is **required** to initialise the virtual filesystem with any previously existing data files in the browser's IndexedDB storage. Without it, the filesystem will always be initially mounted as an empty directory.

For more information, see the Emscripten FS API [`IDBFS` and `FS.syncfs()`](https://emscripten.org/docs/api_reference/Filesystem-API.html#filesystem-api-idbfs) documentation.

### Persisting the filesystem to IndexedDB

The `syncfs()` function should be invoked with its `populate` argument set to `false` to persist the current state of the filesystem to the browser's IndexedDB storage.

::: {.panel-tabset}
## JavaScript

``` javascript
await webR.FS.syncfs(false);
```

## R
```{r eval=FALSE}
webr::syncfs(FALSE)
```

:::

After writing to the virtual filesystem you should be sure to invoke `syncfs(false)` before the web page containing webR is closed to ensure that the filesystem data is flushed and written to the IndexedDB-based persistent storage.

::: {.callout-warning}

Operations performed using IndexedDB are done asynchronously. If you are mounting `IDBFS` filesystems and accessing data non-interactively you should use the JavaScript API and be sure to wait for the `Promise` returned by `webR.FS.syncfs(false)` to resolve before continuing, for example by using the `await` keyword.

In a future version of webR the `webr::syncfs()` function will similarly return a Promise-like object.
:::

### Web storage caveats

Filesystem data stored in an IndexedDB database can only be accessed within the current [origin](https://developer.mozilla.org/en-US/docs/Glossary/Origin), loosely defined as the current web page's host domain and port.

The way in which web browsers decide how much storage space to allocate for data and what to remove when limits are reached differs between browsers and is not always simple to calculate. Be aware of browser [storage quotas and eviction criteria](https://developer.mozilla.org/en-US/docs/Web/API/Storage_API/Storage_quotas_and_eviction_criteria) and note that data stored in an `IDBFS` filesystem type is stored only on a "best-effort" basis. It can be removed by the browser at any time, autonomously or by the user interacting through the browser's UI.

In private browsing mode, for example, stored data is usually deleted when the private session ends.

## JupyterLite DriveFS Filesystem Storage

When using webR as a [JupyterLite kernel](https://github.com/r-wasm/jupyterlite-webr-kernel), the in-browser storage provided by JupyterLite can be mounted using the `DRIVEFS` filesystem type.

The webR kernel mounts the default Jupyterlite storage drive automatically at startup with the mountpoint `/drive`, and so usually you will not need to mount the storage manually.


### Mounting

An optional drive name can be given as the `driveName` option or `source` argument. When omitted, the default value for `source` is the empty string, the default drive name used by JupyterLite.

::: {.panel-tabset}
## JavaScript

``` javascript
await webR.FS.mkdir("/drive");
await webR.FS.mount("DRIVEFS", { driveName: "" }, "/drive");
```

## R

```{r eval=FALSE}
webr::mount(mountpoint = "/drive", source = "", type = "DRIVEFS")
```

:::
