---
title: "R Packages"
format: html
toc: true
---

WebR supports the loading of R packages that have been pre-compiled targeting
WebAssembly and for use with webR.

WebR packages must be installed before use, which in practice means copying
binaries into the webR virtual filesystem in some way, either as part of the
webR build process or during runtime by downloading packages from a repository.

## Downloading packages from a webR binary repository

A collection of supported packages are publicly hosted via a CDN in a CRAN-like
repository with URL,

    https://repo.r-wasm.org/

The public CDN build of webR has been distributed with a pre-installed [webr
support package](rapi). This R package provides a helper function
[`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo),
which can be used to install further R packages from a compatible webR binary
repository.

The repository URL is set using the `repos` argument and the public build of
webR defaults the `repo` argument to the repository URL shown above.

### Example: Installing the Matrix package

Run the following R code in a running webR session, such as the [webR REPL
demo](https://webr.r-wasm.org/latest/),

```{r eval=FALSE}
webr::install("Matrix")
```

If the package is available for webR and downloads successfully, you can then
load the package in the usual way.

```{r eval=FALSE}
library("Matrix")
```

## Interactively installing packages

As part of startup, the [webR REPL demo](https://webr.r-wasm.org/latest/) website
runs the following R code:

```{r eval=FALSE}
webr::global_prompt_install()
```

This function, also provided by the pre-installed [webr support
package](api/r.qmd), installs a global handler that will show an interactive
prompt whenever loading a library fails, asking the user if they would like to
try downloading the package from the configured default webR binary package
repository.

Once the global handler has been installed in this way, it is possible to
install and load packages interactively via a prompt. The process is as follows,

``` r
library(Matrix)
```

    Failed to load package "Matrix". Do you want to try downloading
    it from the webR binary repo?
      1: Yes
      2: No
    Selection: 1
    Downloading webR package: lattice
    Downloading webR package: Matrix

Once the package has been downloaded in this way, the next call to `library()`
will load the package as usual.

## Installing packages from JavaScript

The webR JavaScript API provides the function
[`WebR.installpackages()`](api/js/classes/WebR.WebR.md#installpackages). This
convenience function takes an array of packages as its only argument and for
each calls
[`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo)
with the default webR binary package repository.

``` javascript
await webR.installPackages(['Matrix', 'cli'])
```

Once the promise returned by
[`WebR.installPackages()`](api/js/classes/WebR.WebR.md#installpackages) has
resolved, the packages can be loaded in the usual way using `library()`.

## Building R packages for webR

::: callout-note
It is not possible to install packages from source in webR. This is not likely
to change in the near future, as such a process would require an entire C and
Fortran compiler toolchain to run inside the browser. For the moment, providing
pre-compiled WebAssembly binaries is the only supported way to install R
packages in webR.
:::

Many R packages contain C or C++ code and so must be compiled for WebAssembly
(along with any prerequisite system libraries) before they can be loaded in
webR. One way to do this is by using the
[`Makevars`](https://cran.r-project.org/doc/manuals/r-devel/R-exts.html#Using-Makevars)
system to instruct R to compile R packages with the
[Emscripten](https://emscripten.org/index.html) compiler toolchain, setting
variables such as `CC` and `CXX` to point to the relevant Emscripten compilers.

Once the R package has been compiled and binaries targeting WebAssembly are
available, they can be loaded into webR by providing a CRAN-like repo to the
`webr::install()` command.

More detailed information about the repository and package building process will
be made available in the future, when the infrastructure is more stable.
