---
title: "Installing R Packages"
format: html
toc: true
---

WebR supports the loading of R packages that have been pre-compiled targeting WebAssembly.

WebR packages must be installed before use, which in practice means copying binaries into the webR virtual filesystem in some way, either as part of the webR build process or during runtime by downloading packages from a repository or [mounting filesystem data](packages.qmd#mounting-an-r-library-filesystem-image).

## Downloading packages from a webR binary repository

A collection of supported packages are publicly hosted via a CDN in a CRAN-like repository with URL,

    https://repo.r-wasm.org/

The public CDN build of webR has been distributed with a pre-installed [webr support package](api/r.qmd). This R package provides a helper function [`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo), which can be used to install further R packages from a compatible webR binary repository.

The repository URL is set using the `repos` argument and the public build of webR defaults this argument to the repository URL shown above^[Other R package repositories providing  WebAssembly binaries are also available, such as [R-universe](https://r-universe.dev/).].

### Example: Installing the Matrix package

Run the following R code in a running webR session, such as the [webR REPL demo](https://webr.r-wasm.org/latest/),

```{r eval=FALSE}
webr::install("Matrix")
```

If the package is available for webR and downloads successfully, you can then load the package in the usual way.

```{r eval=FALSE}
library("Matrix")
```

## Shimming `install.packages()` for webR

The [webr support package](api/r.qmd) includes the function [`shim_install()`](api/r.html#shim_install) which optionally replaces specific base R functions with implementations that work in the webR environment. Once executed, the base R `install.packages()` function can be used to install webR packages.

```{r eval=FALSE}
webr::shim_install()
install.packages("Matrix")
```

The [webR REPL demo](https://webr.r-wasm.org/latest/) performs this base R function replacement as part of its startup procedure.

## Interactively installing packages

The `install()` and `require()` base R functions can be shimmed by the function [`shim_install()`](api/r.html#shim_install) so that when a package is missing it is downloaded from the default webR binary package repository.

An interactive prompt may be shown, asking the user if they would like to download the package. To enable the interactive menu, the `webr.show_menu` global option should be set to `TRUE`. Otherwise, the user will not be prompted and the package will be downloaded automatically.

```{r eval=FALSE}
webr::shim_install()
options(webr.show_menu = `TRUE`)
```

Once the menu has been enabled, it is possible to install and load packages interactively via a prompt. The process looks as follows,

``` r
library(Matrix)
```

    Failed to load package "Matrix". Do you want to try downloading
    it from the webR binary repo?
      1: Yes
      2: No
    Selection: 1
    Downloading webR package: lattice
    Downloading webR package: Matrix

After the package has been downloaded and installed, the original `library()` or `require()` function will be invoked to continue loading the package as usual.

### Handling missing packages using a global handler function

An optional global handler is also provided that can be used to automatically download packages if loading fails. The handler can be installed with [`global_prompt_install()`](api/r.html#global_prompt_install) and allows packages to be automatically downloaded without explicitly loading them with `library()` or `require()`. When using the global handler, prompting the user with an interactive menu is also controlled by the `webr.show_menu` global option.

```{r eval=FALSE}
options(webr.show_menu = FALSE)
webr::global_prompt_install()
cli::cli_alert_success("The cli package is working!")
```

    Downloading webR package: cli
    ✔ The cli package is working!

::: callout-note
When the global handler is installed from JavaScript the R code must be evaluated without any other calling handlers on the stack. This can be arranged by setting the `withHandlers` option to `false`^[When evaluating R code in this way R error conditions cannot easily be captured. An alternative is to use [`shim_install()`](api/r.html#shim_install) and explicitly load R packages using `library()` or `require()`.].

``` javascript
await webR.evalR(`
  webr::global_prompt_install()
  cli::cli_alert_success('The cli package works!')
`, { withHandlers: false });
```

:::

## Installing packages from JavaScript

The webR JavaScript API provides the function [`WebR.installPackages()`](api/js/classes/WebR.WebR.md#installpackages). This convenience function takes an array of packages as its only argument and for each calls [`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo) with the default webR binary package repository.

``` javascript
await webR.installPackages(['Matrix', 'cli'])
```

Once the promise returned by [`WebR.installPackages()`](api/js/classes/WebR.WebR.md#installpackages) has resolved, the packages can be loaded in the usual way using `library()`.

## Mounting an R library filesystem image

R libraries that have been packaged with Emscripten's [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool may be loaded into the webR virtual filesystem (VFS) by mounting the image using [`webr::mount()`](api/r.qmd#mount). When using webR in a browser, `type = "WORKERFS"` (the default) should be used so that the filesystem image is downloaded from the URL given by `source` and mounted on the VFS at `mountpoint`.

```{r eval=FALSE}
webr::mount(
  mountpoint = "/library",
  source = "https://example.com/library.data"
)
```

Once the library filesystem image has been mounted, the base R `.libPaths()` function should be used to ensure `mountpoint` is included in R's package library search path.

```{r eval=FALSE}
.libPaths(c(.libPaths(), "/library"))
```

Packages that exist in the R library filesystem image can then be loaded as normal.

```{r eval=FALSE}
library(mypackage)
```

Details on how to build a WebAssembly R package library can be found in the [Building R Packages](building.qmd) section, and further information about mounting data to the VFS (including when running under Node.js) can be found in the [Mounting Filesystem Data](mounting.qmd) section.
