---
title: "Installing R Packages"
format: html
toc: true
---

WebR supports the loading of R packages that have been pre-compiled targeting WebAssembly and for use with webR.

WebR packages must be installed before use, which in practice means copying binaries into the webR virtual filesystem in some way, either as part of the webR build process or during runtime by downloading packages from a repository.

## Downloading packages from a webR binary repository

A collection of supported packages are publicly hosted via a CDN in a CRAN-like repository with URL,

    https://repo.r-wasm.org/

The public CDN build of webR has been distributed with a pre-installed [webr support package](api/r.qmd). This R package provides a helper function [`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo), which can be used to install further R packages from a compatible webR binary repository.

The repository URL is set using the `repos` argument and the public build of webR defaults the `repo` argument to the repository URL shown above^[Other WebAssembly R package repositories are also available, such as [R-universe](https://r-universe.dev/).].

### Example: Installing the Matrix package

Run the following R code in a running webR session, such as the [webR REPL demo](https://webr.r-wasm.org/latest/),

```{r eval=FALSE}
webr::install("Matrix")
```

If the package is available for webR and downloads successfully, you can then load the package in the usual way.

```{r eval=FALSE}
library("Matrix")
```

## Shimming `install.packages()` for webR

The [webr support package](api/r.qmd) includes the function [`shim_install()`](api/r.html#shim_install) which optionally replaces specific base R functions with implementations that work in the webR environment. Once executed, the base R `install.packages()` function can be used to install packages.

```{r eval=FALSE}
webr::shim_install()
install.packages("Matrix")
```

The [webR REPL demo](https://webr.r-wasm.org/latest/) performs this base R function replacement as part of startup.

## Interactively installing packages

As part of startup, the [webR REPL demo](https://webr.r-wasm.org/latest/) website runs the following R code:

```{r eval=FALSE}
webr::global_prompt_install()
```

This function, also provided by the pre-installed [webr support package](api/r.qmd), installs a global handler that will show an interactive prompt whenever loading a library fails, asking the user if they would like to try downloading the package from the configured default webR binary package repository.

Once the global handler has been installed in this way, it is possible to install and load packages interactively via a prompt. The process is as follows,

``` r
library(Matrix)
```

    Failed to load package "Matrix". Do you want to try downloading
    it from the webR binary repo?
      1: Yes
      2: No
    Selection: 1
    Downloading webR package: lattice
    Downloading webR package: Matrix

Once the package has been downloaded in this way, the next call to `library()` will load the package as usual.

## Installing packages from JavaScript

The webR JavaScript API provides the function [`WebR.installPackages()`](api/js/classes/WebR.WebR.md#installpackages). This convenience function takes an array of packages as its only argument and for each calls [`install()`](api/r.qmd#install-one-or-more-packages-from-a-webr-binary-package-repo) with the default webR binary package repository.

``` javascript
await webR.installPackages(['Matrix', 'cli'])
```

Once the promise returned by [`WebR.installPackages()`](api/js/classes/WebR.WebR.md#installpackages) has resolved, the packages can be loaded in the usual way using `library()`.

## Mounting an R library filesystem image

R libraries that have been packaged with Emscripten's [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool may be loaded into the webR virtual filesystem (VFS) by mounting the image using [`webr::mount()`](api/r.qmd#mount). When using webR in a browser, `type = "WORKERFS"` (the default) should be used so that the filesystem image is downloaded from the URL given by `source` and mounted on the VFS at `mountpoint`.

```{r eval=FALSE}
webr::mount(
  mountpoint = "/library",
  source = "https://example.com/library.data"
)
```

Once the library filesystem image has been mounted, the base R `.libPaths()` function should be used to ensure `mountpoint` is included in R's package library search path.

```{r eval=FALSE}
.libPaths(c(.libPaths(), "/library"))
```

Packages that exist in the R library filesystem image can then be loaded as normal.

```{r eval=FALSE}
library(mypackage)
```

Details on how to build a WebAssembly R package library can be found in the [Building R Packages](building.qmd) section, and further information about mounting data to the VFS (including when running under Node.js) can be found in the [Mounting Filesystem Data](mounting.qmd) section.
